---
  html_document:
    toc: true
    highlight: zenburn
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```
### Background
A-G 1-24 Central Rods with HDAC4
H-P 1-24 Peripheral Rods with HDAC4

A-G are cells from one part of the retina and H-P are from cells in a
different part of the retina. We know the peripheral ones are more
resistant to death, but no one knows why that is. So the interesting
biological question is what genes if any are different between the two
populations. For the scrb-seq data, the questions we are asking for
are…


1. Are there any genes or groups of genes with differences between A-G and H-P?
(and if there aren't, is it because there are too few reads per cells/genes detected)

2. Do all of these cells look like rod photoreceptors?
i.e. do they express the appropriate rod genes, and not too many
muller glia, rpe, or cone genes (I attached lists of genes thought to
be enriched in rods [2 different ones], muller glia enriched genes,
rpe enriched genes, cone enriched genes, and rod bipolar cell enriched
genes)

3. How much diversity is there in photoreceptors?  Mostly people have always
envisioned the photoreceptors to be the same, but no one has really looked at
the single cell level before.  If we look at common housekeepers, how much
variability is there?
(I attached a list of housekeepers)

4. Because this was a trial run, I think Magali actually dumped the equivalent
of 384x3 (1152) cells into a single sequencing lane.
Is there any way to know that if we
repeated the experiment, but with fewer cells,
whether the level of detection of the genes would improve?

### Are there groups of genes with differences between A-G and H-P?
First load in the counts and make a heatmap of the correlation between the
two sample groups.

```{r load-counts}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(knitr)
library(reshape)
counts_file = "~/cache/wu-retina-singlecell/counts/scrb-trimarchi/Cepko_Trimarchi_buffer.all.refseq.total.dat"
counts = read.table(counts_file, header=TRUE, sep="\t", row.names=1, stringsAsFactors=FALSE)
```

We can see they kind of cluster into separate groups, with the fovea samples with a higher
correlation with each other.

```{r correlaton-heatmap}
fovea = c("A", "B", "C", "D", "E", "F", "G")
well = unlist(lapply(colnames(counts), function(x) strsplit(x, "_")[[1]][2]))
row = substr(well, 1, 1)
class = unlist(lapply(row, function(x) ifelse(x %in% fovea, "fovea", "periphery")))
class_df = data.frame(class=class)
dat = cor(counts)
rownames(class_df) = rownames(dat)
class_colors = c("yellow", "purple")
names(class_colors) = c("fovea", "periphery")
pheatmap(dat, annotation_colors=list(class=class_colors), annotation=class_df,
    show_rownames=FALSE, show_colnames=FALSE, cluster_rows=FALSE)
pheatmap(dat, annotation_colors=list(class=class_colors), annotation=class_df,
    show_rownames=FALSE, show_colnames=FALSE, cluster_rows=TRUE)
```

These samples have a really low number of counts, about 20k counts per sample. This is
going to make it difficult to do any kind of differential expression analysis.

Papers like this have a million or more: http://www.nature.com/nmeth/journal/v11/n2/extref/nmeth.2772-S1.pdf

```{r counts-per-sample}
qplot(colSums(counts)) + geom_histogram() + scale_x_sqrt() + scale_y_sqrt()
```

Remove genes that are never detected and remove wells with very poor coverage.
```{r differential-expression-setup}
library(DESeq)
library(scde)
counts = counts[rowSums(counts) > 0,]
counts = counts[,colSums(counts)> 10000]
well = unlist(lapply(colnames(counts), function(x) strsplit(x, "_")[[1]][2]))
row = substr(well, 1, 1)
class = unlist(lapply(row, function(x) ifelse(x %in% fovea, "fovea", "periphery")))
class_df = data.frame(class=class)
```


Clustering the samples and genes shows there is not too much of a pattern going on between


```{r gene-heatmap}
pheatmap(sqrt(counts), annotation_colors=list(class=class_colors), annotation=class_df,
    show_rownames=FALSE, show_colnames=FALSE, cluster_rows=FALSE)
```

```{r differential-run}
well = unlist(lapply(colnames(counts), function(x) strsplit(x, "_")[[1]][2]))
row = substr(well, 1, 1)
class = unlist(lapply(row, function(x) ifelse(x %in% fovea, "fovea", "periphery")))
class = as.factor(class)
names(class) = colnames(counts)
class = as.factor(class)
o.ifm = scde.error.models(counts=counts, groups=class, save.crossfit.plots=F,
   n.cores=1, threshold.segmentation=T, save.model.plots=F, verbose=0)
valid.cells <- o.ifm$corr.a >0;
o.ifm = o.ifm[valid.cells,]
counts = counts[, valid.cells]
class = class[valid.cells]
o.prior <- scde.expression.prior(models=o.ifm,counts=counts, length.out=400,show.plot=T)
```

```{r differential-expression, results='asis'}
ediff = scde.expression.difference(o.ifm, counts, o.prior, groups=class, n.randomizations=100,
                                   n.cores=1, verbose=0)
ediff$pvalue = pnorm(-(abs(ediff$Z))) * 2
ediff$padj = p.adjust(ediff$pvalue)
ediff = ediff[order(ediff$pvalue),]
kable(subset(ediff, padj < 0.20), format="markdown")
```

We can find a few differentially expressed genes between the peripheral/fovea cells. The trouble is we see Tsix and Xist, which are two genes that are sex-specific, so we might be looking at sex specific differences here rather than retina/peripheral differences. Xist serves to deactivate the extra X chromosome in females.

```{r xist-plot}
xist = melt(counts[rownames(counts) == "Xist",])
xist$class = class
ggplot(xist, aes(class, value)) + geom_boxplot() + scale_y_sqrt() + xlab("")
```

At any rate we can output a list of the top 100 genes by p-value; just because things aren't
significantly different doesn't mean you can't follow up on hits that you prioritize. It
just means if you wanted to publish this you'd likely have to do some kind of follow up
experiment and not just publish the list, though there can be exceptions.

```{r de-output}
write.table(head(ediff, 100), file="fovea-periphery.tsv", sep="\t", quote=FALSE,
  col.names=TRUE, row.names=TRUE)
```

We can't really tell if there was more reads if we would do a better job calling differentially
expressed genes, we can do it the other way by downsampling when we know we have enough and
look and see where the results fall off, but can't really do the reverse.

The take home of this, though is the mice should be sexxed prior to doing this, because we're
picking up sex specific genes.

### Are these rod photoreceptors?

```{r rod-and-cones}
rod = c("Rho","Sag","Pdc","Gnat1","Pde6g","Rom1","Gngt1","Unc119",
        "Gnb1","Gcap1","Guca1a","Pde6b","Guca2a","Prph2","Aipl1",
        "Tulp1","Rs1","Rcvrn","Cngb1")
roska_cone = c("Pde6c","Opn1sw","Fabp7","Gnat2","Opn1mw","Pde6h","Clca3",
              "Clca3","Clca3","Agr2","Arr3","Gulo","Otop3","Ppm1j","Mogat1","Mogat1",
              "Osgep","Ppp1r14d","Cngb3","Igj","1700113O17Rik","Gngt2","En2","Rbp3",
              "Rbp3","Lcn2","Acbd6")
rod = c(rod, "Pde6b","Nr2e3","Cnga1","Pde6a","A930006D01Rik","Rp1","Cyld","Sh2d1a","Guca1b","Faim","Gm12689","Gm12689","Gnb1","Fscn2","Reep6","Nxnl2","Nxnl1","Nrl","Gm13251","Gm13251","1700020D05Rik","Spata1","Olfr772","Slc24a1","Ybx3","Gm9958","Gm9958","Rex2","Rexo2","Ppap2c","Hmgb2")
col_totals = data.frame(depth=colSums(counts), detected=colSums(counts > 0))
rownames(col_totals) = colnames(counts)
col_totals = col_totals[order(col_totals$depth),]
counts = counts[, rownames(col_totals)]
rod_indexes = rownames(counts) %in% rod
cone_indexes = rownames(counts) %in% roska_cone
```

Below is a heatmap of the rod and cone genes, the samples are on the Y axis and the genes on
the X axis. The units are the square root of the raw counts, to compress onto the scale.

```{r rod-and-cone-heatmap}
anno = data.frame(type=ifelse(rod_indexes & cone_indexes, "both", ifelse(rod_indexes, "rod",
 ifelse(cone_indexes, "cone", "none"))))
rownames(anno) = rownames(counts)

a = subset(anno, type != "none")

pheatmap(t(sqrt(counts[cone_indexes | rod_indexes,])), show_colnames=TRUE, cluster_rows=TRUE, annotation=a, show_rownames=FALSE)
```

Hard to tell if these are cones or rods; there are some cone markers that are expressed
at a pretty good clip but there is a set of rod markers enriched as well.


### Diversity in photoreceptors?
I don't think the variability isn't going to tell you too much,
because there is variability just due to whether or not the mRNA are
captured for sequencing at all with single cell RNA-seq. Below is a
heatmap of the housekeeping genes, ordered by the total number of
counts for the sample. You can see the number of genes
detected correlates the overall number of reads in each cell, and that
in turn has a correlation to the depth of the housekeeping
genes. The presence/absence of the housekeeping genes says more about
the variation in sequencing than the samples than the content of the cells.

First a replot of the cone/rod genes using normalized counts so we can compare.

Below is the same
heatmap, but using the normalized counts, since we use that below to look at glia/other
enrichment:

```{r rod-heatmap}
housekeeping = c("Prpf4b","Ndfip1","Gpx1","Polr2a","Rps27a","Rpl13a","Mapk6","Eif2s1","Sin3b","Tfr2","Polr2g","Ywhae","Rps3","Tbp","Gyk","Hmbs","Vim","Pfkp","Aldoa","Rpl11","Nono","Cdca8","Tfrc","Gm13105","Alas1","Sbk1","Hprt","Arhgdia","Gapdh","Slc25a12","Ubc","Eef1a1","Dvl1","Ctxn1","G6pd2","Hint1","Rps15","Tpm2","Tubb4a")
pheatmap(sqrt(counts[rod_indexes | cone_indexes,]), show_colnames=FALSE, cluster_rows=TRUE, show_rownames=TRUE, cluster_cols=FALSE, annotation=col_totals)
```

And now of the housekeeping genes:

```{r norm-housekeeping-heatmap}
pheatmap(sqrt(counts[rownames(counts) %in% housekeeping,]), show_colnames=FALSE, cluster_rows=TRUE, show_rownames=TRUE, annotation=col_totals, drop_levels=FALSE, cluster_cols=FALSE)
```

#### Is there more variation in the photoreceptor genes compared to the housekeeping genes?

As a dirty way of checking it, calculate the CV for the photoreceptor and housekeeping genes,
then do a t.test.

```{r cv-test}
photo_cv = apply(counts[cone_indexes | rod_indexes,], 1, sd) / apply(counts[cone_indexes | rod_indexes,], 1, mean)
house_cv = apply(counts[housekeeping,], 1, sd) / apply(counts[housekeeping,], 1, mean)
t.test(photo_cv, house_cv)
```

#### Is there any glial/rpe cell contamination?
```{r glial-rpe-contamination}
glia=c("Apoe","Aqp4","Clu","Vim","Kcnj2","Kcnj12","Kcnj10","S100a16","Glul","Rlbp1","Dkk3","Vsx2","Spc25","Itm2b","Atg4b","Dbi","Tspo","Gpr37","Car2","Abca8a","Sox2","Gnai2","Cav2","Car5a","Cav1","Jagn1","Dapl1","Synpr","Col9a1","Gm5771","Prss3","2210010C04Rik","Gm10334","Prss1","Prss2","Uba3","Gnb1l","Wdr46","Kcnj10","","Ctsh","Rhpn1")
rpe = c("Rpe65","Ttr","Crx","Dct","Best1","Six3","Chrna3","Trpm1","Lhx2","Sfrp5","Pmel","Cspg5","Aplp1","Pdxp","Arid4a","Rbp1","Tyrp1","Myrip","Tfpi2","Mab21l1","Ptgds","Frzb","Slc6a15","Serpinf1","Dusp4","Gmnn","Gem","Bmp4","Slc6a20a","Slc6a20b","Cdo1","Ttll4","Cdh15","Cdh3","Aldh1a3","Clcn4-2","Enpp2","Col8a2","Gpr143","Sostdc1","Pdpn","Gpnmb","Sorbs2","Sulf1","Efemp1","Foxd1","Lnx1","Mpdz","Sdc2","Gja1","Limch1","Fgfr2","Plcb4","Gas1")
```

It's hard to tell if there is contamination, we don't see a lot of these markers expressed,
barring a couple outliers, which is good.

```{r glial-contamination}
pheatmap(sqrt(counts[rownames(counts) %in% glia,]), show_colnames=FALSE, cluster_rows=TRUE, show_rownames=TRUE, annotation=col_totals, drop_levels=FALSE, cluster_cols=FALSE)
```

```{r rpe-contamination}
pheatmap(sqrt(counts[rownames(counts) %in% rpe,]), show_colnames=FALSE, cluster_rows=TRUE, show_rownames=TRUE, annotation=col_totals, drop_levels=FALSE, cluster_cols=FALSE)
```


#### What are the genes with the highest variation?

We fit a line to the mean vs coefficient of variation plot and find the 100 genes
which deviate the most from the calculated CV. These are genes that are much more
highly variable than expected.

```{r cv-fit}
require(DESeq)
lib.size <- estimateSizeFactorsForMatrix(counts)
ed <- t(t(counts)/lib.size)

vars <- apply(ed,1,var)
means <- rowMeans(ed)
cv2 <- vars/means^2
require(statmod)
minMeanForFit <- unname( quantile( means[ which( cv2 > .3 ) ], .95 ) )
useForFit <- means >= minMeanForFit # & spikeins
fit <- glmgam.fit( cbind( a0 = 1, a1tilde = 1/means[useForFit] ),cv2[useForFit] )
a0 <- unname( fit$coefficients["a0"] )
a1 <- unname( fit$coefficients["a1tilde"])
fit$coefficients
par(mar=c(3.5,3.5,1,1),mgp=c(2,0.65,0),cex=0.9); smoothScatter(log(means),log(cv2));
xg <- exp(seq( min(log(means[means>0])), max(log(means)), length.out=1000 ))
vfit <- a1/xg + a0
# add fit line
df <- ncol(ed) - 1
# add confidence interval
afit <- a1/means+a0
varFitRatio <- vars/(afit*means^2)
varorder <- order(varFitRatio,decreasing=T)
oed <- ed[varorder,]
par(mar=c(3.5,3.5,1,1),mgp=c(2,0.65,0),cex=0.9); smoothScatter(log(means),log(cv2)); lines( log(xg), log(vfit), col="black", lwd=3 ); lines(log(xg),log(vfit * qchisq(0.975,df)/df),lty=2,col="black"); lines(log(xg),log(vfit * qchisq(0.025,df)/df),lty=2,col="black");
# add top 100 genes
points(log(means[varorder[1:100]]),log(cv2[varorder[1:100]]),col=2)
```

This identifies a set of genes that are more variable than expected:

```{r variance-set}
pval <- pchisq(varFitRatio*df,df=df,lower.tail=F)
adj.pval <- p.adjust(pval,"fdr")
sigVariedGenes <- adj.pval<1e-3;
table(sigVariedGenes)
sig_list = adj.pval[sigVariedGenes & !is.na(sigVariedGenes)]
sig_varied = data.frame(padj=sig_list)
rownames(sig_varied) = names(sig_list)
write.table(sig_varied, file="significant_variation.tsv", sep="\t",
 col.names=TRUE, row.names=TRUE, quote=FALSE)
```

Are any of these photoreceptor or housekeeping genes?
```{r photoreceptor-and-housekeeping-variance}
photoreceptor = c(rod, roska_cone)
sig_varied$id = rownames(sig_varied)
```

photoreceptors:
```{r photoreceptor-variance, results='asis'}
photoreceptor_varied = sig_varied[rownames(sig_varied) %in% photoreceptor,]
kable(sig_varied[rownames(sig_varied) %in% photoreceptor,], format='markdown')
```

housekeeping:
```{r housekeeping-variance, results='asis'}
housekeeping_varied = sig_varied[rownames(sig_varied) %in% housekeeping,]
kable(sig_varied[rownames(sig_varied) %in% housekeeping,], format='markdown')
```

Are these driven by a small number of outlier samples or is there just more overall
variation for the photoreceptor genes?

```{r housekeeping-variance-heatmap}
pheatmap(sqrt(counts[rownames(counts) %in% photoreceptor_varied$id,]), show_colnames=FALSE, cluster_rows=TRUE, show_rownames=TRUE, annotation=col_totals, drop_levels=FALSE)
```

It looks like there are a small number of samples that have high expression of these genes.

### Wrapup

1. We identified some DE genes, but it looks like there might be a gender bias that is
   swamping the signal.
2. It is up in the air if these look like rod photoreceptors or not, there are some cone
   markers expressed and photoreceptor markers expressed in the samples.
3. The primary driver of variability is overall number of reads and number of genes detected.
If we normalize the counts to total number of reads, we still see that the number of
reads and genes detected correlates to the expression of housekeeping genes. We did some
dirty measures of measuring variation and found that there might be a couple of photoreceptor
genes that might vary between individual cells, but the low count rate makes it hard to
know for sure.
4. We can't answer if sequencing deeper would help. We only pick up at most 3.5k genes,
and for some samples even less than that, so for many samples we're not seeing the
genes at all. Is that due to incomplete capturing of the mRNA or not sequencing deep
enough? We can't really tell.

While the new SCRB-seq is better than the old one, it is still not picking out very
many genes. If we look at the number of genes detected per sample, it is still very low,
around 1000 genes.

```{r genes-detected-histogram}
qplot(colSums(counts > 0)) + geom_histogram() + scale_x_sqrt() + scale_y_sqrt() +
   xlab("genes detected") + ylab("cells")
```

Some of the alternate methods seem to be detecting a lot more than
that, but I have no personal experience with those datasets yet. For example:

Picelli, S., Björklund, A. K., Faridani, O. R., Sagasser, S., Winberg, G., & Sandberg, R. (2013). Smart-seq2 for sensitive full-length transcriptome profiling in single cells. Nature Methods, 10(11), 1096–1098. doi:10.1038/nmeth.2639

Having only 20,000 counts per cell on average is killer as well. There also
seems to be a large number of the reads concentrated in a single gene, Malat1:

```{r counts-per-gene}
 qplot(rowSums(counts)) + geom_histogram() + scale_x_sqrt() + scale_y_sqrt()
```

I'm not sure what is so special about Malat1 that it is found so consistently in every
sample. I'm suspecting there is some kind of technical issue with it:

```{r malat1}
rowSums(counts)[rowSums(counts) > 4e05]
summary(rowSums(counts))
```
